### 2022-06-17 draft.tex
Deploying to gh-pages from @ PoisotLab/ms_betacov_mapping@fa58cdffbc410db83637c21cff76d58ca3bd6ac3 🚀

\fancypagestyle{normal}
{
  \fancyhf{}
  \fancyfoot[R]{\footnotesize\sffamily\thepage\ of \pageref*{LastPage}}
}
\begin{document}
\raggedright
\thispagestyle{empty}
{\Large\bfseries\sffamily The coevolutionary mosaic of
bat-betacoronaviruses spillover risk}
\vskip 5em

@@

Spillover risk is multidimensional and complicated. Within a pool of
competent hosts, it is driven by a multiplicity of factors (Plowright et
al. 2017). Although proxies for the local richness of hosts is commonly
analysed (see \emph{e.g.} Anthony et al. 2017 for coronaviruses), there
is an argument to be made that species who are not (known as) competent
hosts of a specific virus genus may not factor into this (Plowright et
al. 2015), calling for species-level information. This is especially
true as competence data increases predictive power when the taxonomic
scope of hosts of a viral family increases (Becker et al. 2020, Mull et
al. 2022). Similarly, host species who share viruses at different rates
should be weighted accordingly (Albery et al. 2020). In mammals, key
functional traits (for which phylogeny is a reasonable proxy) are
determinants of the spillover potential (Olival et al. 2017); these
include, notably, body mass, and affinity for urban environments (Albery
et al. 2022). Finally, especially when the pool of potential hosts spans
the entire globe, there may be local host pools that are highly unique;
not having been observed in other locations, these can act on the
overall risk either by providing novel contact opportunities, reflecting
unique host-environment combinations (Engering et al. 2013), or
facilitating rapid evolutionary changes in specialism of their pathogens
(Agosta et al. 2010). In the specific case of generalist pathogens
(which is the case many viruses in the betacoronavirus genus, see
\emph{e.g.} MacLean et al. 2021), there is conceptual and empirical
support to the idea that these community-level mechanisms are even more
important in driving the overall risk (Power and Mitchell 2004).

@@

We identify the Amazon and South-Eastern Asia as hotspots where the
phylogenetic distinctiveness of betacoronaviruses is the highest
(Anthony et al. 2017); surprisingly, current data suggest that viral
sharing between hosts is high in the Amazon and low in South-Eastern
Asia, which has the potential to result in different evolutionary
dynamics between these two regions, hinting at different futures for
their viral communities. This work is important both as a description of
the bats-betacoronavirus complex, but also because more broadly, bats
are known reservoirs for a variety of emerging viruses and other
pathogens (Calisher et al. 2006, Melaun et al. 2014), making balancing
the needs for bat conservation and disease prevention most likely very
difficult and a source of human-wildlife conflicts, especially in more
densely populated areas (Stone et al. 2015, Rego et al. 2015).

@@

\hypertarget{known-betacoronavirus-hosts}{%
\subsection{Known betacoronavirus
hosts}\label{known-betacoronavirus-hosts}}

We downloaded the data on bats hosts of betacoronaviruses assembled by
Becker et al. (2022) from
\texttt{https://www.viralemergence.org/betacov} on Apr.~2022, and
filtered it to ``known'' hosts (established before the emergence of
SARS-CoV-2) and ``novel'' hosts (confirmed through sampling since the
emergence of SARS-CoV-2). The original database was assembled by a
combination of data mining and literature surveys, including automated
alerts on the ``bats'' and ``coronavirus'' keywords to identify novel
empirical evidence of bats-betacoronaviruses associations.

@@

We downloaded the rangemap of every extant bat species that was either
classified as an empirically documented host of beta-coronaviruses from
the previous step, according to recent IUCN data (IUCN 2021). The range
maps were subsequently rasterized using the \texttt{rasterize} function
from \texttt{GDAL} (Rouault et al. 2022) at a resolution of
approximately \textbf{TK TP}. For every pixel in the resulting raster
where at least one bat host of betacoronavirus was present, we extract
the species pool (list of all bat species), which was used to calculate
the following risk assessment components: phylogenetic diversity, bat
compositional uniqueness, and predicted viral sharing risk.

\hypertarget{bats-phylogeography}{%
\subsection{Bats phylogeography}\label{bats-phylogeography}}

@@

For all bat hosts of betacoronaviruses, we extracted their predicted
viral sharing network (Albery et al. 2020). This network stores pairwise
values of viral community similarity. To project viral sharing values
into a single value for every pixel, we averaged the pairwise scores.
High values of the average sharing propensity means that this specific
extant bat assemblage is likely to be proficient at exchanging viruses.

@@

We used the following query to pull all betacoronavirus sequence data
from the GenBank Nucleotide database except SARS-CoV-2;
(``Betacoronavirus''{[}Organism{]} OR betacoronavirus{[}All Fields{]})
NOT (``Severe acute respiratory syndrome coronavirus 2''{[}Organism{]}
OR sars-cov-2{[}All Fields{]}). We added a single representative
sequence for SARS-CoV-2 and manually curated to remove sequences without
the RNA-dependent RNA polymerase (RdRp) sequence or that contained words
indicating recombinant or laboratory strains including ``patent,''
``mutant,'' ``GFP,'' and ``recombinant.'' We filtered over-represented
taxa including betacoronavirus 1, hCoV-OC43, Middle East respiratory
syndrome coronavirus, Murine hepatitis virus, and hCoV-HKU1. Curated
betacoronavirus RdRp sequences were then aligned using MAFFT v 1.4.0
(\textbf{Katoh and Standley 2013}, parameters in text?) and a maximum
likelihood tree reconstructed in IQ-TREE v 1.6.12 (Nguyen et al.~2015)
with ModelFinder (Kalyaanamoorthy et al.~2017) ultrafast bootstrap
approximation (Hoang et al. 2018) and the following parameters
(\textbf{STEPH WILL ADD}, parameters in text?).

@@

Subsequently, we tested the hypothesis that the biogeography of bat
betacoronaviruses should track the biogeography of their hosts. To test
this idea, we loosely adapted a method from (Kreft and Jetz 2007, 2010),
who proposed a phylogenetic method for the delineation of animal
biogeographic regions. In their original method, a distance matrix -
where each row or column represents a geographic raster's grid cell, and
the dissimilarity values are the ``beta diversity similarity'' of their
community assemble - undergoes non-metric multidimensional scaling
(NMDS); the first two axes of the NMDS are projected geographically
using a four-color bivariate map. Here, we build on this idea with an
entirely novel methodology. First, we measure the phylogenetic distance
between the different viruses in the betacoronavirus tree by using the
cophenetic function in \texttt{ape} (Paradis and Schliep 2019);
subsequently, we take a principal components analysis of that distance
matrix (readily interchangeable for NMDS in this case) to project the
viral tree into an n-dimensional space. We then take the first two
principal components and, as with the evolutionary distinctiveness
analysis, aggregated these to a mean host value and projected them using
a four-color bivariate map.

@@

Bats are found worldwide and are both one of the most diverse groups
among mammals and one of the main animal reservoir for different strains
of betacoronaviruses (Drexler et al. 2014). This has attracted attention
to areas where high diversity of bats, and therefore presumably high
diversity of betacoronaviruses, can be an important issue for human
health (Calisher et al. 2006, Moratelli and Calisher 2015). By
overlaying the IUCN rangempas for confirmed bat hosts of
betacoronaviruses {[}fig.~\ref{fig:richness}; top{]}, we see that the
the main hotspots of host richness are primarily South-Eastern Asia,
parts of Southern Europe, and to a lesser extent parts of Africa in the
-25-0 range of latitudes. The description of host richness is an
important first step towards understanding risk, as previous research
(Anthony et al. 2017, Mollentze and Streicker 2020) states that locally
diverse bat communities could maintain more viruses and hence, a higher
probability of having a pathogen that could represent a risk for human
health.

\begin{figure}
\hypertarget{fig:richness}{%
\centering
\includegraphics{figures/combined_richness.png}
\caption{Top panel: relative diversity of known bat hosts of
betacoronaviruses. This map shows that the region with the largest
number of possible hosts is South-Eastern Asia. Bottom panel: congruence
between the evolutionary distinctiveness of the hosts (grey to blue) and
the viruses (grey to red). By contrast to the richness map, this reveals
that South America has the most evolutionary distinct hosts \emph{and}
viruses, whereas South-Eastern Asia has mostly distinct viruses. This is
congruent with know results about New World bats being evolutionary
distinct, and suggests that they similarly have distinct
viruses.}\label{fig:richness}
}
\end{figure}

Nevertheless, locally diverse and virus-rich bat communities could
represent an increased risk of spillover under climate change through
the creation of novel interactions (Carlson et al. 2022), and therefore
the diversity of betacoronavirus strains should similarly be ccounted
for. In fig.~\ref{fig:richness} (bottom), we contrast the evolutionary
distinctiveness of bats and viruses -- this reveals a slightly different
portrait than bat richness alone. Chiropterans can be classified, from a
macro-evolutionary standpoint, as microchiroptera and macrochiroptera,
where macrochiroptera have an older history from an evolutionary
perspective compared to macrochiroptera (Teeling et al. 2005, Springer
2013). Specifically, we would expect that the so-called ``New World''
group of bats, being more evolutionary distinct, would also have
evolutionary distinct viruses. Indeed fig.~\ref{fig:richness} (bottom)
reveals it to be the case, and this region harbors a distinct
bat-betacoronavirus complex. By contrast, South-Eastern Asia has a lot
of non-evolutionary distinct bats, but evolutionary-distinct viruses.

It is noteworthy that outside of South America, viral evolutionary
distinctiveness does not accurately tracks host diversity, with some
areas having over-distinct viruses (southern China but, oddly, not the
rest of southeast Asia). There are a number of likely explanations.
First, given the richness of bats in southeast Asia, many
betacoronaviruses likely remain to be discovered in this region. Indeed,
global predictions by Becker et al. (2022) highlight that southeast Asia
is a likely hostpot of unconfirmed hosts of betacoronaviruses, which
would likely result in additiona viral discoveries. This idea is
unsurprising given the growing realization, around the emergence of
SARS-CoV-2, that a unique lineage of similar viruses are widespread in
bats but still mostly undescribed. The most distinct
bats/betacoronavirus complex is found in South America, a region with a
comparatively lower number of hosts; this matches with the isolation
through viariance of the host group, and may highlight a different
co-evolutionary dynamic. Alternatively, this distinctiveness hostpot may
be a product of under-sampling: South-America is one of the places where
the fewest betacoronaviruses have been discovered (Anthony et al. 2017,
Olival et al. 2017, Allen et al. 2017), resulting in sparser
phylogenetic tree, thereby artificially inflating distinctiveness.
Adding more viruses would bring the distinctiveness of known sequences
down.

@@

Despite the difference in evolutionary distinctiveness globally, there
are reasons to expect that the phylogeography of bats and
betacoronaviruses should show some degree of congruence. High density of
hosts sharing the same virus (albeit possibly different strains) can
drive or result from evolution of the bat antiviral immune system,
resulting in spatially distinct immunological responses, as evidenced in
several bat species (Banerjee et al. 2020). Immune characteristics that
allow bats to be better adapted to infection by emerging viruses
(Gorbunova et al. 2020, Irving et al. 2021), in addition to being
hardcoded in their genome (Jebb et al. 2020), may be related to a wide
variety of diets (Banerjee et al. 2020, Moreno Santillán et al. 2021,
Muylaert et al. 2021), themselves likely to be driven by spatial
effects, especially at the local scale -- bats, indeed, occupy a variety
of environments, and therefore display a variety of adaptations to these
environments.

@@

In fig.~\ref{fig:biogeo}, we show a projection of the phylogeographic
signal of bats (top) and viruses (bottom) in space; the distinct
groupings (represented by different colors symbolizing positions in the
subspace formed by the first two axes of the PCoA) are essentially
equivalent between the two groups, and can be coarsely delineated as
southeast Asia, Eurasia above a northing of 25, and Africa and south
America. These results suggest that, although the evolutionary
distinctiveness of the bat/betacoronavirus complex varies spatially, the
system shows an important degree of spatial consistency, with a reduced
number of bioregions. Available information describing the spillover of
zoonotic betacoronaviruses of bat origin where data was available before
and up through the COVID-19 pandemic puts spillover events of SARS-CoV-2
in Wuhan, China; SARS-CoV in Guangdong, China based on the presence of
closest known viruses circulating in nature, and a nearby location where
serological (antibody) evidence has indicated human exposure to
SARS-like viruses (Wang et al. 2018); MERS-CoV in Saudi Arabia based on
index cases available from a recently-published compendium of cases
(Ramshaw et al. 2019). For the latest event, most if not all index cases
are presumed to be camel-to-human transmission, and the precise origin
point (if it exists) of MERS-CoV in bats is uncertain. Recent
recombinant canine coronavirus spillover events in Haiti (Lednicky et
al. 2021) and Europe (Vlasova et al. 2022) are not relevant here, as
bats' involvement in these cycles of transmission have been supposed to
be non-existent. These index cases fall within different phylogeographic
bioregions (fig.~\ref{fig:biogeo}), which further highlight the issue
that different host-virus sub-systems may lead to widespread emergence.

@@

As host richness, joint disctinveness, or phylogeographic structure
suggest that the bat/betacoronavirus complex is globally fragmented
enough to give rise to both different levels of risk (as evidenced by
the spatial location of spillover events) and different types of
co-evolutionary dynamics, we turn to the Geographic Mosaic Theory of
Coevolution (Thompson 2005) to provide a measure of risk accounting for
multiple processes. In fig.~\ref{fig:trivariate}, we overlapped three
components of spillover risk: viral sharing, \emph{i.e.} the chance that
two bats will share viruses overall; Local Contribution to Beta
Diversity, \emph{i.e.} the fact that a bat community is compositionally
unique compared to the average compositional similarity across the
entire system; finally, host phylogenetic diversity, \emph{i.e.} how
dispersed the bats in a location are within the tree of life. This
approach leads to the definition of broad biogeographic regions of risk,
where the same color represents the same type of risk. By way of
constrat to figures fig.~\ref{fig:richness} and fig.~\ref{fig:biogeo},
these regions do not necessarilly overlap with previous spatial
partitions of the bat/betacoronavirus complex.

\begin{figure}
\hypertarget{fig:trivariate}{%
\centering
\includegraphics{figures/risk_trivariate.png}
\caption{Trivariate additive mapping of the components of risk in the
red/green/blue, where high virus sharing is encoded in the blue channel,
host phylogenetic diversity in the green channel, and compositional
uniqueness in the red channel. A pixel that would maximize all measures
would be a pure white (specifically RGB(1.0, 1.0. 1.0)), and a pixel
with the lowest possible values would be pure black (specifically
RGB(0.0, 0.0, 0.0)). Therefore, lighter values (the sum of the three
channels gets closer to 3) indicate higher risk, and the color indicates
the proportional distribution of the factors making up the total
risk.}\label{fig:trivariate}
}
\end{figure}

From the perspective of spillover risk, the most important combination
of factors is a high phylogenetic diversity of hosts with low viral
sharing; this, essentially, means that very different betacoronavirus
could co-exist within the same place. This is particularly the case
given that betacoronaviruses often evolve and even achieve host shifts
through recombination, which requires the co-occurrence of sufficiently
distinct viruses to be a major driver of emergence. In
fig.~\ref{fig:trivariate}, this corresponds to yellow to pale green
areas, which are essentially limited to South-Eastern Asia, and to some
part of Sub-Saharan Africa. Adopting a geographic mosaic theory
perspective on risk, other regions of the world are of lesser concern
fig.~\ref{fig:risk}. Our risk decomposition does not account for viral
diversity or distinctiveness. The simple rationale behind it is that the
acquisition of viral data is rarely disconnected from the acquisition of
host data; furthermore, there are more sources of information on hosts
than on viruses, allowing to develop a host-centric perspective on risk.
Areas with high bat diversity and high turnover \emph{may} facilitate
the evolutionary radiation of viruses, matching previous findings that
the diversification of bat coronaviruses is driven largely by host
shifts (inter-genus or higher levels of cross-species transmission) and,
to a lesser degree, cospeciation and sharing (intra-genus cross-species
transmission; Anthony et al. 2017). This diversification is not an
actual risk factor for spillover itself, but acts downstream of a
spillover event by increasing the random chance of the emergence of a
virus with the raw genomic components required for the potential to
infect humans.

\begin{figure}
\hypertarget{fig:risk}{%
\centering
\includegraphics{figures/risk_map.png}
\caption{Extraction of a measure of risk based on the colorimetric space
from fig.~\ref{fig:trivariate}. The risk is a composite measure of the
color value and angular distance to the yellow hue, as defined in the
methods, ranged in the unit space. Based on this analyses, regions at
high risk of spillover are southeast Asia and
Madagascar.}\label{fig:risk}
}
\end{figure}

From another perspective, areas of high host uniqueness and virus
sharing (red-to-pink) could provide hotspots of betacoronavirus risk
through mixing of unique viruses (via codivergence) and in turn
recombination. Under our framework, such a hotspot was identified in
Madagascar, where most bat species are endemic following evolutionary
divergence from sister species in both African and Asian continents
(\emph{e.g.} Shi et al. 2014). Recent surveillance (Kettenburg et al.
2022) has identified a novel betacoronavirus (in the subgenus
\emph{Nobecovirus}) in Madagascar-endemic pteropid bat species
(\emph{Pteropus rufus}, \emph{Rousettus madagascariensis}), emphasizing
strong proof of principle in model predictions.

@@

@@

>>>>

\fancypagestyle{normal}
{
  \fancyhf{}
  \fancyfoot[R]{\footnotesize\sffamily\thepage\ of \pageref*{LastPage}}
}
\begin{document}
\raggedright
\thispagestyle{empty}
{\Large\bfseries\sffamily The coevolutionary mosaic of
\emph{Betacoronavirus} spillover risk from bat hosts}
\vskip 5em

@@

Spillover risk is multidimensional and complicated. Within a pool of
competent hosts, it is driven by a multiplicity of factors (Plowright et
al. 2017). Although proxies for the local richness of hosts is commonly
analysed (see \emph{e.g.} Anthony et al. 2017 for coronaviruses), there
is an argument to be made that species who are not (known as) competent
hosts of a specific virus genus may not factor into this (Plowright et
al. 2015), calling for species-level information. This is especially
true as competence data increases predictive power when the taxonomic
scope of hosts of a viral family increases (Becker et al. 2020, Mull et
al. 2022). Similarly, host species who share viruses at different rates
should be weighted accordingly (Albery et al. 2020). In mammals, key
functional traits (for which phylogeny is a reasonable proxy) are
determinants of the spillover potential (Olival et al. 2017); these
include, notably, body mass, and affinity for urban environments (Albery
et al. 2022). Finally, especially when the pool of potential hosts spans
the entire globe, there may be local host pools that are highly unique;
not having been observed in other locations, these can act on the
overall risk either by providing novel contact opportunities, reflecting
unique host-environment combinations (Engering et al. 2013), or
facilitating rapid evolutionary changes in specialism of their pathogens
(Agosta et al. 2010). In the specific case of generalist pathogens
(which is the case many viruses in the \emph{Betacoronavirus} genus, see
\emph{e.g.} MacLean et al. 2021), there is conceptual and empirical
support to the idea that these community-level mechanisms are even more
important in driving the overall risk (Power and Mitchell 2004).

@@

We identify the Amazon and South-Eastern Asia as hotspots where the
phylogenetic distinctiveness of \emph{Betacoronavirus} is the highest
(Anthony et al. 2017); surprisingly, current data suggest that viral
sharing between hosts is high in the Amazon and low in South-Eastern
Asia, which has the potential to result in different evolutionary
dynamics between these two regions, hinting at different futures for
their viral communities. This work is important both as a description of
the bats-betacoronaviruses complex, but also because more broadly, bats
are known reservoirs for a variety of emerging viruses and other
pathogens (Calisher et al. 2006, Melaun et al. 2014), making balancing
the needs for bat conservation and disease prevention most likely very
difficult and a source of human-wildlife conflicts, especially in more
densely populated areas (Stone et al. 2015, Rego et al. 2015).

@@

\hypertarget{known-betacoronavirus-hosts}{%
\subsection{\texorpdfstring{Known \emph{Betacoronavirus}
hosts}{Known Betacoronavirus hosts}}\label{known-betacoronavirus-hosts}}

We downloaded the data on bats hosts of \emph{Betacoronavirus} assembled
by Becker et al. (2022) from
\texttt{https://www.viralemergence.org/betacov} on Apr.~2022, and
filtered it to ``known'' hosts (established before the emergence of
SARS-CoV-2) and ``novel'' hosts (confirmed through sampling and
competence assays since the emergence of SARS-CoV-2). The original
database was assembled by a combination of data mining and literature
surveys, including automated alerts on the ``bats'' and ``coronavirus''
keywords to identify novel empirical evidence of bats-betacoronaviruses
associations; this yielded a total of 126 known hosts, 47 of which were
novel hosts.

@@

We downloaded the rangemap of every current bat species that was
classified as an empirically documented host of \emph{Betacoronavirus}
from the previous step, according to recent IUCN data (IUCN 2021). The
range maps were subsequently rasterized using the \texttt{rasterize}
function from \texttt{GDAL} (Rouault et al. 2022) at a resolution of
approximately 100kmx100km. For every pixel in the resulting raster where
at least one bat host of \emph{Betacoronavirus} was present, we extract
the species pool (list of all competent bat hosts), which was used to
calculate the following risk assessment components: bat phylogenetic
diversity, bat compositional uniqueness, and predicted viral sharing
risk.

\hypertarget{bats-phylogenetic-diversity}{%
\subsection{Bats phylogenetic
diversity}\label{bats-phylogenetic-diversity}}

@@

For all bat hosts of \emph{Betacoronavirus}, we extracted their
predicted viral sharing network, generated from a previously published
generalized additive mixed model of virus sharing by a tensor function
of phylogenetic distance and geographic range overlap (Albery et al.
2020). This network stores pairwise values of viral community
similarity. To project viral sharing values into a single value for
every pixel, we averaged the pairwise scores. High values of the average
sharing propensity means that this specific extant bat assemblage is
likely to be proficient at exchanging viruses.

@@

To next represent phylogeography of betacoronaviruses in bats, we
aggregated and analyzed betacoronavirus sequence data. We used the
following query to pull all \emph{Betacoronavirus} sequence data from
the GenBank Nucleotide database except SARS-CoV-2;
(``Betacoronavirus''{[}Organism{]} OR betacoronavirus{[}All Fields{]})
NOT (``Severe acute respiratory syndrome coronavirus 2''{[}Organism{]}
OR sars-cov-2{[}All Fields{]}). We added a single representative
sequence for SARS-CoV-2 and manually curated to remove sequences without
the RNA-dependent RNA polymerase (RdRp) sequence or that contained words
indicating recombinant or laboratory strains including ``patent,''
``mutant,'' ``GFP,'' and ``recombinant.'' We filtered over-represented
taxa including betacoronavirus 1, hCoV-OC43, Middle East respiratory
syndrome coronavirus, Murine hepatitis virus, and hCoV-HKU1. Curated
betacoronavirus RdRp sequences were then aligned using MAFFT v 1.4.0
{[}Katoh and Standley (2013); Algorithm FFT-NS-2, Scoring matrix 200PAM
/ k=2, gap open penalty 1.53m offset value 0.123{]} and a maximum
likelihood tree reconstructed in IQ-TREE v 1.6.12 (Nguyen et al. 2015)
with ModelFinder (Kalyaanamoorthy et al. 2017) ultrafast bootstrap
approximation (Hoang et al. 2018) with a general time reversible model
with empirical base frequencies and the 5-discrete-rate-category
FreeRaye model of nucleotide substitution (GTR+F+R5).

@@

Subsequently, we tested the hypothesis that the biogeography of bat
betacoronaviruses should track the biogeography of their hosts. To test
this idea, we loosely adapted a method from (Kreft and Jetz 2007, 2010),
who proposed a phylogenetic method for the delineation of animal
biogeographic regions. In their original method, a distance matrix -
where each row or column represents a geographic raster's grid cell, and
the dissimilarity values are the ``beta diversity similarity'' of their
community assemble - undergoes non-metric multidimensional scaling
(NMDS); the first two axes of the NMDS are projected geographically
using a four-color bivariate map. Here, we build on this idea with an
entirely novel methodology. First, we measure the phylogenetic distance
between the different viruses in the betacoronaviruses tree by using the
cophenetic function in \texttt{ape} (Paradis and Schliep 2019);
subsequently, we take a principal components analysis of that distance
matrix (readily interchangeable for NMDS in this case) to project the
viral tree into an n-dimensional space. We then take the first two
principal components and, as with the evolutionary distinctiveness
analysis, aggregated these to a mean host value and projected them using
a four-color bivariate map.

@@

Bats, the second most diverse groups of mammals, are found worldwide and
serve as the main animal reservoir for different strains of
betacoronaviruses (Drexler et al. 2014). This has attracted attention to
areas where high diversity of bats, and therefore presumably high
diversity of betacoronaviruses, can be an important issue for human
health (Calisher et al. 2006, Moratelli and Calisher 2015). By
overlaying the IUCN range maps for confirmed bat hosts of
betacoronaviruses {[}fig.~\ref{fig:richness}; top{]}, we see that the
the main hotspots (both in terms of size and higher values) of host
richness are primarily South-Eastern Asia, parts of Southern Europe, and
to a lesser extent parts of Africa in the -25-0 range of latitudes. The
description of host richness is an important first step towards
understanding risk, as previous research (Anthony et al. 2017, Mollentze
and Streicker 2020) states that locally diverse bat communities could
maintain more viruses and hence, a higher probability of having a
pathogen that could represent a risk for human health.

\begin{figure}
\hypertarget{fig:richness}{%
\centering
\includegraphics{figures/combined_richness.png}
\caption{Top panel: relative diversity of known bat hosts of
betacoronaviruses. This map shows that the region with the largest
number of possible hosts is South-Eastern Asia. Bottom panel: congruence
between the evolutionary distinctiveness of the hosts (grey to blue) and
the viruses (grey to red). By contrast to the richness map, this reveals
that South America has the most evolutionary distinct hosts \emph{and}
viruses, whereas South-Eastern Asia and the Rift Valley region have
mostly distinct viruses. This is congruent with know results about New
World bats being evolutionary distinct, and suggests that they similarly
have distinct viruses.}\label{fig:richness}
}
\end{figure}

Nevertheless, locally diverse and virus-rich bat communities could
represent an increased risk of spillover under climate change through
the creation of novel interactions (Carlson et al. 2022), and therefore
the diversity of \emph{Betacoronavirus} strains should similarly be
ccounted for. In fig.~\ref{fig:richness} (bottom), we contrast the
evolutionary distinctiveness of bats and viruses -- this reveals a
slightly different portrait than bat richness alone. Chiropterans can be
classified, from a macro-evolutionary standpoint, as Yangochiroptera in
the ``New World,'' and Yinpterochiroptera elsewhere (Teeling et al.
2005, Springer 2013). Specifically, we would expect that the so-called
``New World'' group of bats, being more evolutionary distinct, would
also have evolutionary distinct viruses. Indeed fig.~\ref{fig:richness}
(bottom) reveals it to be the case, and this region harbors a distinct
bat-betacoronaviruses complex. By contrast, South-Eastern Asia has a lot
of non-evolutionary distinct bats, but evolutionary-distinct viruses.

It is noteworthy that outside of South America, viral evolutionary
distinctiveness does not accurately track host diversity, with some
areas having over-distinct viruses (eastern China but, oddly, not the
rest of southeast Asia). There are a number of likely explanations.
First, given the richness of bats in southeast Asia, many
betacoronaviruses likely remain to be discovered in this region. Indeed,
global predictions by Becker et al. (2022) highlight that southeast Asia
is a likely hostpot of unconfirmed hosts of betacoronaviruses, which
would likely result in additional viral discoveries. This idea is
unsurprising given the growing realization, especially since the
emergence of SARS-CoV-2, that unique lineages of similar viruses are
widespread in bats but still mostly undescribed. The most distinct
bats-betacoronaviruses complex is found in South America, a region with
a comparatively lower number of hosts; this matches with the isolation
through viariance of the host group, and may highlight a different
co-evolutionary dynamic. Alternatively, this distinctiveness hostpot may
be a product of under-sampling: South-America is one of the places where
the fewest \emph{Betacoronavirus} sequences have been discovered
(Anthony et al. 2017, Olival et al. 2017, Allen et al. 2017), resulting
in sparser phylogenetic tree, thereby artificially inflating
distinctiveness. Adding more viruses would bring the distinctiveness of
known sequences down. Finally, South-America is the range of
Phyllostomidae, a group of bats that underwent explosive diversification
events (Villalobos and Arita 2010), which may drive the emergence of
multiple viral lineages.

@@

Despite the difference in evolutionary distinctiveness globally, there
are reasons to expect that the phylogeography of bats and
betacoronaviruses should show some degree of congruence (Letko et al.
2020, Van Brussel and Holmes 2022). In particular, this should be the
case if viruses can circulate among hosts and co-evolve with local hosts
communities, making their evolutionary process more than a byproduct of
host evolution. High density of hosts sharing the same virus (albeit
possibly different strains) can drive or result from evolution of the
bat antiviral immune system, resulting in spatially distinct
immunological responses, as evidenced in several bat species (Banerjee
et al. 2020). Immune characteristics that allow bats to be better
adapted to infection by emerging viruses (Gorbunova et al. 2020, Irving
et al. 2021), in addition to being hardcoded in their genome (Jebb et
al. 2020), may be related to a wide variety of diets (Banerjee et al.
2020, Moreno Santillán et al. 2021), themselves likely to be driven by
spatial effects, especially at the local scale -- bats, indeed, occupy a
variety of environments, and therefore display a variety of adaptations
to these environments (Muylaert et al. 2022).

@@

In fig.~\ref{fig:biogeo}, we show a projection of the phylogeographic
signal of bats (top) and viruses (bottom) in space; the distinct
groupings (represented by different colors symbolizing positions in the
subspace formed by the first two axes of the PCoA) are essentially
equivalent between the two groups, and can be coarsely delineated as
southeast Asia, Eurasia above a northing of 25, and Africa and south
America. These results suggest that, although the evolutionary
distinctiveness of the bat-betacoronaviruses complex varies spatially,
the system shows an important degree of spatial consistency, with a
reduced number of bioregions. Available information describing the
spillover of zoonotic betacoronaviruses of bat origin where data was
available before and up through the COVID-19 pandemic puts spillover
events of SARS-CoV-2 in Wuhan, China; SARS-CoV in Guangdong, China based
on the presence of closest known viruses circulating in nature, and a
nearby location where serological (antibody) evidence has indicated
human exposure to SARS-like viruses (Wang et al. 2018); MERS-CoV in
Saudi Arabia based on index cases available from a recently-published
compendium of cases (Ramshaw et al. 2019). For the latest event, most if
not all index cases are presumed to be camel-to-human transmission, and
the precise origin point (if it exists) of MERS-CoV in bats is
uncertain. Recent recombinant canine coronavirus spillover events in
Haiti (Lednicky et al. 2021) and Europe (Vlasova et al. 2022) are not
relevant here, as bats' involvement in these cycles of transmission have
been supposed to be non-existent. These index cases fall within
different phylogeographic bioregions (fig.~\ref{fig:biogeo}), which
further highlight the issue that different host-virus sub-systems may
lead to widespread emergence.

@@

As host richness, joint distinctiveness, or phylogeographic structure
suggest that the bat-betacoronaviruses complex is globally fragmented
enough to give rise to both different levels of risk (as evidenced by
the spatial location of spillover events) and different types of
co-evolutionary dynamics, we turn to the Geographic Mosaic Theory of
Coevolution (Thompson 2005) to provide a measure of risk accounting for
multiple processes. In fig.~\ref{fig:trivariate}, we overlapped three
components of spillover risk: viral sharing, \emph{i.e.} the chance that
two bats will share viruses overall; Local Contribution to Beta
Diversity, \emph{i.e.} the fact that a bat community is compositionally
unique compared to the average compositional similarity across the
entire system; finally, host phylogenetic diversity, \emph{i.e.} how
dispersed the bats in a location are within the tree of life. This
approach leads to the definition of broad biogeographic regions of risk,
where the same color represents the same type of risk. By way of
constrat to figures fig.~\ref{fig:richness} and fig.~\ref{fig:biogeo},
these regions do not necessarilly overlap with previous spatial
partitions of the bat-betacoronaviruses complex.

\begin{figure}
\hypertarget{fig:trivariate}{%
\centering
\includegraphics{figures/risk_trivariate.png}
\caption{Trivariate additive mapping of the components of risk in the
red/green/blue, where high virus sharing is encoded in the blue channel,
host phylogenetic diversity in the green channel, and compositional
uniqueness in the red channel. A pixel that would maximize all measures
(highest possible risk) would be a pure white (specifically RGB(1.0,
1.0. 1.0)), and a pixel with the lowest possible values would be pure
black (specifically RGB(0.0, 0.0, 0.0)). Therefore, lighter values (the
sum of the three channels gets closer to 3) indicate higher risk, and
the color indicates the proportional distribution of the factors making
up the total risk.}\label{fig:trivariate}
}
\end{figure}

From the perspective of spillover risk, the most important combination
of factors is a high phylogenetic diversity of hosts with low viral
sharing; this, essentially, means that very different betacoronaviruses
could co-exist within the same place. This is particularly the case
given that betacoronaviruses often evolve and even achieve host shifts
through recombination, which requires the co-occurrence of sufficiently
distinct viruses to be a major driver of emergence. In
fig.~\ref{fig:trivariate}, this corresponds to yellow to pale green
areas, which are essentially limited to South-Eastern Asia, and to some
part of Sub-Saharan Africa. Adopting a geographic mosaic theory
perspective on risk, other regions of the world are of lesser concern
fig.~\ref{fig:risk}. Our risk decomposition does not account for viral
diversity or distinctiveness. The simple rationale behind it is that the
acquisition of viral data is rarely disconnected from the acquisition of
host data. There are more sources of information on hosts than on
viruses, allowing to develop a host-centric perspective on risk
(although this estimate would more accurate with viral traits related to
\emph{e.g.} ability to switch hosts or pathogenic potential). Areas with
high bat diversity and high turnover \emph{may} facilitate the
evolutionary radiation of viruses, matching previous findings that the
diversification of bat coronaviruses is driven largely by host shifts
(inter-genus or higher levels of cross-species transmission) and, to a
lesser degree, cospeciation and sharing (intra-genus cross-species
transmission; Anthony et al. 2017). This diversification is not an
actual risk factor for spillover itself, but acts downstream of a
spillover event by increasing the random chance of the emergence of a
virus with the raw genomic components required for the potential to
infect humans.

\begin{figure}
\hypertarget{fig:risk}{%
\centering
\includegraphics{figures/risk_map.png}
\caption{Extraction of a measure of \emph{Betacoronavirus} spillover
risk from bat hosts based on the colorimetric space from
fig.~\ref{fig:trivariate}. The risk is a composite measure of the color
value and angular distance to the yellow hue, as defined in the methods,
ranged in the unit space. Based on this analyses, regions at high risk
of spillover are southeast Asia and Madagascar.}\label{fig:risk}
}
\end{figure}

From another perspective, areas of high host uniqueness and virus
sharing (red-to-pink) could provide hotspots of \emph{Betacoronavirus}
risk through mixing of unique viruses (via codivergence) and in turn
recombination. Under our framework, such a hotspot was identified in
Madagascar, where most bat species are endemic following evolutionary
divergence from sister species in both African and Asian continents
(\emph{e.g.} Shi et al. 2014). Recent surveillance (Kettenburg et al.
2022) has identified a novel \emph{Betacoronavirus} (in the subgenus
\emph{Nobecovirus}) in Madagascar-endemic pteropid bat species
(\emph{Pteropus rufus}, \emph{Rousettus madagascariensis}), emphasizing
strong proof of principle in model predictions.

@@

\leavevmode\hypertarget{ref-Cohen2022SamStr}{}%
Cohen, L. E. et al. 2022. Sampling strategies and pre-pandemic
surveillance gaps for bat coronaviruses.: 2022.06.15.496296.

@@

\leavevmode\hypertarget{ref-Hoang2018UfbImp}{}%
Hoang, D. T. et al. 2018. UFBoot2: Improving the Ultrafast Bootstrap
Approximation. - Molecular Biology and Evolution 35: 518--522.
